---
title: "Bird Density Patterns"
author: "Nicole Barker"
output:
  html_document: default
  word_document:
    reference_docx: ../RmdStyles/ReportFormat_1.docx
---

# =================
# UNFINISHED STUFF
# change colour palette for height 
# REDO AGGREGATES OF RANKS SO THAT VALUES FOR UNRANKED COMBINATIONS EXIST WHEN I USE GGPLOT FUNCTION (i.e. fake data solution. )
# =================



**Objective: Summarize how observed bird density varies across forest attributes**

We need to convert observed Counts to predicted Density only to correct for survey method and detectability. We can't describe patterns in counts because it is confounded with different contributed datasets. 

**Desired result**: A density estimate for each forest stand type that has at least one point count survey

**Stats hold up: I'm not sure of the best way to summarize this information, given the nested design of our survey. I'm worried about pseudoreplication.** 

**Concerns:**

* Multiple surveys for a given location (visits within PKEY; PKEYs per SS)
* Multiple locations for a given stand (SS per ObjectID)
* Multiple stands sampled for a given stand type (ObjectID for ForId)

I think I should accommodate this structure when I calculate summary statistics, figures, etc. But I'm not sure how.

*As of August 20 2018, I'm a bit worried about whether I'm running the right analyses. But for the purposese of cleaning up my scripts, I'm going to continue forth with the headings and code. I will have a meeting with Lionel (and others) at some point and update these scripts with decisions made based on how to adjust for pseudoreplication.*

``` {r setup, echo=F, message=F, warning=F}
require(knitr)
opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(
  fig.path = "figures/"
)
rm(list=ls())
getwd()
```

``` {r}
source("Rfunctions/UsefulFunctions.R")
```


#### Method: Poisson GLM with 
* Forest stand type as predictor
* Count as response variable
* TOTA offset as offset 

Count + offset ~ ForID

#### Dataset requirements

1. mastermatrix
2. ranks
3. bird observations
4. QPAD offsets

#### Data subset:

Those stand types that were sampled by at least one bird point count survey.

## Step 1. Load data

####1. mastermatrix

Contains all possible forest stand type combinations, based on factorial combination of all levels of all 4 attributes. The combination of forest attributes is the unique ForID for a stand type.

```{r load.matrixTemplate, echo=T, eval=T}
load("RDataFiles/mastermatrix.RData")
mastermatrix$ForID.qs <- mastermatrix$ForID
mastermatrix$ForID <- as.character(mastermatrix$ForID)
kable(head(mastermatrix, 10), row.names = F)
matrix.qs <- mastermatrix
```

####2. ranks

Contains stand ranks assigned to ForID. Note that ranks were previously quality-checked. These are the final ranks. 

``` {r}
ranks <- read.csv("rawData/ForID_Ranks_final_long.csv", head=T)
ranks$ForID.qs <- ranks$ForID
ranks$ForID <- as.character(ranks$ForID)
ranks.qs <- ranks

kable(t(table(ranks$FinalRank)))
```

As seen above, this file contains text-based ranks that won't work in a GLM. Convert to numeric, switching all text-based ranks to NA. The "0 or 3" is complicated and not worth explaining. We just convert it to NA, indicating 'unranked.

``` {r}
ranks$FinalRank <- as.numeric(as.character(ranks$FinalRank))

kable(head(ranks, 10))
kable(t(table(ranks$FinalRank)))
```


####3. bird observations

Bird data from BAM avian dataset available for sampled forest stands:

1) SS: sampling location; point count location within a forest stand
2) PKEY: unique survey identity
3) Count: sum of birds per survey (PKEY)
4) ForID: unique forst stand type within which the SS is located


```{r load.pointcountloc.data, echo=T, eval=T}
load("RDataFiles/totalBirds.RData")
totalBirds$ForID.qs <- totalBirds$ForID
totalBirds$ForID <- as.character(totalBirds$ForID)
qs.birds <- totalBirds

totalBirds$SS <- as.character(totalBirds$SS) # convert SS (survey location) to character

kable(rbind(head(totalBirds, 10), tail(totalBirds, 10)))
```


####4. QPAD offsets

Offsets developed using QPAD approach, for "total species". Corrects for survey methodology but not individual species' singing rates or EDRs. Treats all species as one highly-variable species. 


``` {r}
load("RDataFiles/TOTA_offsets-v3_2016-12-01.RData")
qs.off <- OFF

kable(head(OFF, 10))

OFF <- as.data.frame(OFF)
OFF$PKEY <- row.names(OFF)
kable(head(OFF, 10), row.names=F)

```

####5. Previously aggregated datasets

These summaries of average rank across various combinations of forest attribute were generated in D.01.RankPatterns.Rmd

``` {r}
load("RDataFiles/summarizedRanks.RData")
```

## Step 2. Combine datasets and switch column classes as necessary

#### Associate offset with PKEY
``` {r}
totalBirds <- merge(totalBirds, OFF, by="PKEY", all.x=T) # merge by PKEY; keep only those PKEY in bird dataset.
```

#### Associate rank with ForID
``` {r}
mastermatrix <- merge(mastermatrix, ranks, by="ForID", all=T) # merge by ForID. Keep all ForIDs for now (even unranked)
```

#### Associate bird data with ForID
``` {r}
totalBirds <- merge(totalBirds, mastermatrix, by="ForID", all.x=T) # merge by ForID. Keep only sampled ForIDs (i.e., those with survey data)
totalBirds <- totalBirds[c("ForID", "BEC", "SpeciesGroup", "ForAge", "ForHt", "SS", "PKEY", "Count", "TOTA", "FinalRank")]

kable(head(totalBirds, 20), row.names=F)
```

#### Clean up and switch classes as necessary

``` {r}
unique(totalBirds$ForAge) # already an ordered factor
levels(totalBirds$ForAge) <- c("3-30 yrs", "31-80 yrs", "81-120 yrs", "121-250 yrs", ">250 yrs")
unique(totalBirds$ForAge) # already an ordered factor

unique(totalBirds$ForHt) # already an ordered factor
levels(totalBirds$ForHt) <- c("<10.5 m", "10.5-19.4 m", "19.5-28.4 m", "> 28.5 m")
unique(totalBirds$ForHt) # already an ordered factor

totalBirds$BEC <- factor(totalBirds$BEC) # eliminates missing BEC levels
unique(totalBirds$BEC)

totalBirds$SpeciesGroup <- factor(totalBirds$SpeciesGroup) # eliminates missing speciesgroup levels
unique(totalBirds$SpeciesGroup)

levels(mastermatrix$ForAge) <- c("3-30 yrs", "31-80 yrs", "81-120 yrs", "121-250 yrs", ">250 yrs")
levels(mastermatrix$ForHt) <- c("<10.5 m", "10.5-19.4 m", "19.5-28.4 m", "> 28.5 m")
mastermatrix$BEC <- factor(mastermatrix$BEC) # eliminates missing BEC levels
mastermatrix$SpeciesGroup <- factor(mastermatrix$SpeciesGroup) # eliminates missing speciesgroup levels
```

There are `r length(unique(totalBirds$ForID))` sampled forest stand types going into this analysis. 

## Step 3. Calculate mean density per ForID, rather than count 

If we want to plot anything to interpret our results, we can't just plot count beause it is not corrected for method/detectability. 

So here we're just calculating the mean bird density per ForID (forest stand type).

Forest stand type is a factor with `r length(unique(totalBirds$ForID))` levels

**May adjust this based on pseudoreplication discussion above. e.g., using individual stand polygon (not stand TYPE) as a random factor**

``` {r}
load("RDataFiles/fit.dens.forid_tot.RData") # loads the saved results if it's already run and saved

if (!exists("fit.dens.forid_tot")==TRUE) { #conditional statement; won't run if the file was already loaded from RData
  fit.dens.forid_tot <- glm(Count ~ ForID-1, family=poisson, data=totalBirds, offset=TOTA)  # run if not already run and saved
  save(fit.dens.forid_tot, file="RDataFiles/fit.dens.forid_tot.RData") # save it so i don't need to run again. 
}

```

#### Prediction: Estimate the number of birds in each forest stand type, based on the model. 

Only predicts to the forest stands we sampled. `r length(unique(totalBirds$ForID))` unique stand types.
Basically just takes the mean of all samples within a forest stand type and adjusts by the offset. All stands of the same type will have the same predicted bird density. No covariates in the above model.

``` {r}
newdat <- totalBirds[c("ForID")]
newdat$TOTA <- 0 # set offset to zero so the prediction is density, not observed count
totalBirds$PredDens_ForID <- predict(fit.dens.forid_tot, newdata=newdat, type="response")
```

##### Double-check that these predictions make sense. 

Expectation: All point counts within the same forest stand type should have the same expected density, if I understand this analysis correctly. 

``` {r}
testset <- totalBirds[totalBirds$ForID %in% "SBS_SBPS.SwB_Decid.31_80.2",]
unique(testset$PredDens_ForID)
unique(testset$Count)
testset <- totalBirds[totalBirds$ForID %in% "IDF.SwB.121_250.4+",]
unique(testset$PredDens_ForID)
unique(testset$Count)
```

Expectation is true for those example Forest stand types, despite their observed counts being different. 


# Plot Patterns: Visualize relationships between age/height and bird density (in sampled stands)

*As of August 20 2018, I'm a bit worried about whether I'm running the right analyses. Right now this function plots a density diagram of all rows in the totalBirds file. I think each row corresponds to an individual PKEY. There can be more than one PKEY in a given forest stand polygon.  Nevermind the fact that some stand TYPES (forID) have been sampled more than others. My question is about what unit should be treated as the individual replicate for purposes of displaying relationships: The bird point survey (no), the survey location (no?), the individual forest stand (maybe?) or the stand TYPE (maybe?). I'm going to meet with Lione and others to decide which way to present the information. Then I'll need to take a subset of this table and redo the plots below.*


#### Density as a function of rank 

``` {r}
ggplot(totalBirds, aes(x=as.factor(FinalRank), y=PredDens_ForID, group=as.factor(FinalRank), fill=as.factor(FinalRank))) + geom_boxplot() + xlab("Rank") + ylab("Mean Density per Forest Stand Type") + myfacettheme2 + guides(fill=FALSE)

ggplot(totalBirds, aes(x=as.factor(FinalRank), y=PredDens_ForID, group=as.factor(FinalRank), col=as.factor(FinalRank), fill=as.factor(FinalRank))) + geom_violin() + xlab("Rank") + ylab("Mean Density per Forest Stand Type") + myfacettheme2 + guides(col=FALSE, fill=FALSE)
```

#### Absolute count as a function of rank

Disclaimer that there may be patterns here related to sampling rather than any forest attribute. e.g., if projects that used the same survey method also tended to sample the same forest types, we have a systematic bias. 

``` {r}
ggplot(totalBirds, aes(x=as.factor(FinalRank), y=Count, group=as.factor(FinalRank), fill=as.factor(FinalRank))) + geom_boxplot() + xlab("Rank") +  myfacettheme2 + guides(fill=FALSE)

ggplot(totalBirds, aes(x=as.factor(FinalRank), y=Count, group=as.factor(FinalRank), col=as.factor(FinalRank), fill=as.factor(FinalRank))) + geom_violin() + xlab("Rank")  + myfacettheme2 + guides(fill=FALSE, col=FALSE)
```


#### Bird density as a function of forest height and age

Across all BEC and tree species groups.
Boxplot and violin plot show the same information. 
Labelled points show the mean rank across stands of that combination of age and height. 

##### Create fake data to help with plotting

Some combinations of forest attributes were not sampled. When plotting box plots, the box width varies when certain categories don't have data. 

The [solution](https://stackoverflow.com/questions/15367762/include-space-for-missing-factor-level-used-in-fill-aesthetics-in-geom-boxplot) is to create fake data for the combination of factors for which data is currently missing. The data needs to be outside the plotting range of real data. 

First we need to identify which combinations of Forest age and Froest height have missing values. 
Then we create fake data for those combinations. Needs to have the same columns as the data.frame I'll be using to plot (in this case, totalBirds).

``` {r}
agehtcombos.possible <- unique(paste(mastermatrix$ForAge, mastermatrix$ForHt, sep="_and_"))
agehtcombos.sampled <- unique(paste(totalBirds$ForAge, totalBirds$ForHt, sep="_and_"))
agehtcombos.missing <- agehtcombos.possible[!agehtcombos.possible %in% agehtcombos.sampled]

dat2 <- data.frame(str_split_fixed(agehtcombos.missing, "_and_", 2))  # create dataframe for ForID with no bird data
colnames(dat2) <- c("ForAge", "ForHt")
dat4 <- data.frame(ForID=rep(-999, times=nrow(dat2)), BEC=rep(unique(totalBirds$BEC)[1], times=nrow(dat2)), SpeciesGroup=rep(unique(totalBirds$SpeciesGroup)[1], times=nrow(dat2)), ForAge=dat2$ForAge, ForHt=dat2$ForHt, Count=rep(-999, times=nrow(dat2)), FinalRank=rep(-999, times=nrow(dat2)), PredDens_ForID=rep(-999, times=nrow(dat2)))   # create fake data for Count, FinalRank, and PredDens_ForID

test <- rbind(totalBirds[colnames(dat4)], dat4) # combine real data with fake data
```



``` {r eval=T}
ggplot(test, aes(x=ForAge, y=PredDens_ForID, fill=ForHt)) + geom_boxplot(na.rm=FALSE) + xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2 + coord_cartesian(ylim=range(totalBirds$PredDens_ForID) + c(-1, 1)) + geom_point(data=rank.aggregates[[1]], aes(x=ForAge, y=FinalRank+20, col=ForHt), position=position_dodge(width=.75), size=4, show.legend = F) + geom_text(data=rank.aggregates[[1]], aes(x=ForAge, y=FinalRank+21, label=round(FinalRank,1)), position=position_dodge(width=0.75), size=3.1) 

ggplot(test, aes(x=ForAge, y=PredDens_ForID, fill=ForHt)) + geom_boxplot(na.rm=FALSE) + xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2 + coord_cartesian(ylim=range(totalBirds$PredDens_ForID) + c(-1, 1))

ggplot(totalBirds, aes(x=ForAge, y=PredDens_ForID, fill=ForHt, col=ForHt)) + geom_violin() + xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2
```


#### Absolute count as a function of forest height and age

Across all BEC and tree species groups.
Boxplot and violin plot show the same information. 

``` {r eval=T}
ggplot(totalBirds, aes(x=ForAge, y=Count, fill=ForHt)) + geom_boxplot() + xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2

ggplot(totalBirds, aes(x=ForAge, y=Count, fill=ForHt, col=ForHt)) + geom_violin() + xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2
```

``` {r}
rm(test, agehtcombos.missing, agehtcombos.possible, agehtcombos.sampled)
```

### BEC-specific relationships 

Depict pattern in bird density as a function of forest age and height, separated by different BECs.

##### Create fake data to help with plotting

``` {r}
agehtbeccombos.possible <- unique(paste(mastermatrix$ForAge, mastermatrix$ForHt, mastermatrix$BEC, sep="_and_"))
agehtbeccombos.sampled <- unique(paste(totalBirds$ForAge, totalBirds$ForHt, totalBirds$BEC, sep="_and_"))
agehtbeccombos.missing <- agehtbeccombos.possible[!agehtbeccombos.possible %in% agehtbeccombos.sampled]

dat2 <- data.frame(str_split_fixed(agehtbeccombos.missing, "_and_", 3))  # create dataframe for ForID with no bird data
colnames(dat2) <- c("ForAge", "ForHt", "BEC")
dat4 <- data.frame(ForID=rep(-999, times=nrow(dat2)), BEC=dat2$BEC, SpeciesGroup=rep(unique(totalBirds$SpeciesGroup)[1], times=nrow(dat2)), ForAge=dat2$ForAge, ForHt=dat2$ForHt, Count=rep(-999, times=nrow(dat2)), FinalRank=rep(-999, times=nrow(dat2)), PredDens_ForID=rep(-999, times=nrow(dat2)))   # create fake data for Count, FinalRank, and PredDens_ForID

test <- rbind(totalBirds[colnames(dat4)], dat4) # combine real data with fake data
```


``` {r eval=T}
tota.bec <- split(test, test$BEC)
rank.agg.bec <- split(rank.aggregates[[2]], rank.aggregates[[2]]$BEC)

lapply(1:length(tota.bec), function(i) {
  ggplot(tota.bec[[i]], aes(x=ForAge, y=PredDens_ForID, fill=ForHt)) + geom_boxplot() + facet_wrap(~BEC) + xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2 + coord_cartesian(ylim=range(totalBirds$PredDens_ForID) + c(-1, 1))  
    ggplot(tota.bec[[i]], aes(x=ForAge, y=PredDens_ForID, fill=ForHt)) + geom_boxplot() + facet_wrap(~BEC) + xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2 + coord_cartesian(ylim=range(totalBirds$PredDens_ForID) + c(-1, 1)) + geom_point(data=rank.agg.bec[[i]], aes(x=ForAge, y=FinalRank+20, col=ForHt), position=position_dodge(width=.75), size=4, show.legend = F) + geom_text(data=rank.agg.bec[[i]], aes(x=ForAge, y=FinalRank+21, label=round(FinalRank,1)), position=position_dodge(width=0.75), size=3.1) 
})
  
lapply(tota.bec, function(x) {
  ggplot(x, aes(x=ForAge, y=PredDens_ForID, col=ForHt, fill=ForHt)) + geom_violin()  + facet_wrap(~BEC)+ xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2
})
```

``` {r}
rm(test, agehtbeccombos.missing, agehtbeccombos.possible, agehtbeccombos.sampled, tota.bec)
```


### Patterns for different leading tree species groups 

Depict pattern in bird density as a function of forest age and height, separated by different leading tree species groups

##### Create fake data to help with plotting

``` {r}
agehtsgcombos.possible <- unique(paste(mastermatrix$ForAge, mastermatrix$ForHt, mastermatrix$SpeciesGroup, sep="_and_"))
agehtsgcombos.sampled <- unique(paste(totalBirds$ForAge, totalBirds$ForHt, totalBirds$SpeciesGroup, sep="_and_"))
agehtsgcombos.missing <- agehtsgcombos.possible[!agehtsgcombos.possible %in% agehtsgcombos.sampled]

dat2 <- data.frame(str_split_fixed(agehtsgcombos.missing, "_and_", 3))  # create dataframe for ForID with no bird data
colnames(dat2) <- c("ForAge", "ForHt", "SpeciesGroup")

dat4 <- data.frame(ForID=rep(-999, times=nrow(dat2)), BEC=rep(unique(totalBirds$BEC)[1], times=nrow(dat2)), SpeciesGroup=dat2$SpeciesGroup, ForAge=dat2$ForAge, ForHt=dat2$ForHt, Count=rep(-999, times=nrow(dat2)), FinalRank=rep(-999, times=nrow(dat2)), PredDens_ForID=rep(-999, times=nrow(dat2)))   # create fake data for Count, FinalRank, and PredDens_ForID

test <- rbind(totalBirds[colnames(dat4)], dat4) # combine real data with fake data
```


``` {r eval=T}
tota.sg <- split(test, test$SpeciesGroup)

lapply(tota.sg, function(x) {
  ggplot(x, aes(x=ForAge, y=PredDens_ForID, fill=ForHt)) + geom_boxplot() + facet_wrap(~SpeciesGroup) + xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2 + coord_cartesian(ylim=range(totalBirds$PredDens_ForID) + c(-1, 1))  
    ggplot(x, aes(x=ForAge, y=PredDens_ForID, fill=ForHt)) + geom_boxplot() + facet_wrap(~SpeciesGroup) + xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2 + coord_cartesian(ylim=range(totalBirds$PredDens_ForID) + c(-1, 1)) + geom_point(data=rank.aggregates[[1]], aes(x=ForAge, y=FinalRank+20, col=ForHt), position=position_dodge(width=.75), size=4, show.legend = F) + geom_text(data=rank.aggregates[[1]], aes(x=ForAge, y=FinalRank+21, label=round(FinalRank,1)), position=position_dodge(width=0.75), size=3.1) 
})
  
lapply(tota.bec, function(x) {
  ggplot(x, aes(x=ForAge, y=PredDens_ForID, col=ForHt, fill=ForHt)) + geom_violin()  + facet_wrap(~SpeciesGroup)+ xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2
})
```

### Patterns for different combinations of BECs and leading tree species groups 

##### Create fake data to help with plotting

``` {r}
forid.possible <- unique(paste(mastermatrix$ForAge, mastermatrix$ForHt, mastermatrix$BEC, mastermatrix$SpeciesGroup, sep="_and_"))
forid.sampled <- unique(paste(totalBirds$ForAge, totalBirds$ForHt, totalBirds$BEC, totalBirds$SpeciesGroup, sep="_and_"))
forid.missing <- agehtsgcombos.possible[!agehtsgcombos.possible %in% agehtsgcombos.sampled]

dat2 <- data.frame(str_split_fixed(agehtsgcombos.missing, "_and_", 3))  # create dataframe for ForID with no bird data
colnames(dat2) <- c("ForAge", "ForHt", "SpeciesGroup")

dat4 <- data.frame(ForID=rep(-999, times=nrow(dat2)), BEC=rep(unique(totalBirds$BEC)[1], times=nrow(dat2)), SpeciesGroup=dat2$SpeciesGroup, ForAge=dat2$ForAge, ForHt=dat2$ForHt, Count=rep(-999, times=nrow(dat2)), FinalRank=rep(-999, times=nrow(dat2)), PredDens_ForID=rep(-999, times=nrow(dat2)))   # create fake data for Count, FinalRank, and PredDens_ForID

test <- rbind(totalBirds[colnames(dat4)], dat4) # combine real data with fake data
```



``` {r, eval=T}
totalBirds$BEC.Species <- with(totalBirds, paste(BEC, SpeciesGroup, sep="."))
tota.becsg <- split(totalBirds, totalBirds$BEC.Species)

lapply(tota.becsg, function(x) {
  ggplot(x, aes(x=ForAge, y=PredDens_ForID, fill=ForHt, col=ForHt, group=ForHt)) + geom_point(size=3, position=position_dodge(width=0.8)) + facet_wrap(~BEC.Species)+ xlab("Forest Age") + ylab("Mean Density per Forest Stand Type") + myfacettheme2
})
  
```


