---
title: ""
author: "Nicole Barker"
date: "Last run: Feb 1, 2018"
output: 
  word_document:
    reference_docx: ../styles/ReportFormat_1.docx
---
## Script Abstract

One of a series of scripts that quality-checks, corrects,  pre-processes, and merges the various tables from BAM's Avian Database. 

This script 

**SCRIPT OUTPUTS:**

1. **

## Background
On Nov 30, 2017, Trish provided me with the Access Database of BAM's avian data: COFI_BC_NOV30_2017.accdb. Three tables relate to survey method, one to behaviour, 

**FILES**


**OFFSET-REQUIREMENTS**

* To calculate the Solymos et al 2013 style offsets (correcting for survey method, species availability, species perceptability), we need the following details: 
    * XY coordinates (from SS table)
    * time zone (from SS table)
    * tree cover (tree; TREE) at point count location (from SS table)
    * land cover class (NALCMS05) at point count location (from SS table)
    * land cover class (LCC2; LCC4) at point count location (derived from NALCMS)
    * Maximum Duration (MaxDuration; MAXDUR) of survey (from method table)
    * Maximum Distance (Masdist; MAXDIS) of survey (from method table)
    * Date of year (YYYY MM and DD) (from PKEY table)
    * Days since local spring (DSLS; derived from????
    * Time since sunrise (TSSR; derived from survey time andtime zone)
    * species (SPP) (from point count data table)
    
    PKEY
    TSSR
    JDAY
    MAXDUR
    MAXDIS
    
    TREE
    TREE3
    HAB_NALC1
    HAB_NALC2
    SPRING
    MDT_offset (derived from time zone)
    srise (derived from sunriset function and XY coordinates)
    DSLS (derived from JULIAN & SPRING)
    
    

This script does the following

* 
* 
* 

``` {r setup, echo=F, message=F, warning=F}
require(knitr)
opts_knit$set(root.dir = '..')
```


``` {r, warning=F, message=F}
source("lib/Setup_NB_WorkPC.R")
source("lib/Setup_anyComputer.R")
```

