---
title: "A.0X. Prep bird data"
author: "Lionel Leston & Nicole Barker"
date: "Feb 2019"
output: html_document
---
```{r setup, echo=F, message=F, warning=F}
require(knitr)
opts_knit$set(root.dir = '..')
knitr::opts_chunk$set(
  fig.path = "figures/"
)
rm(list=ls()); gc()
getwd()
```

``` {r usefulfunc, echo=F}
source("Rfunctions/UsefulFunctions.R")
```

``` {r}
birds.ss <- read.csv("rawData/BBS_BAM_Atlas_MigLayer_Jan22_2017.csv", header=T)
colnames(birds.ss)
unique(birds.ss$PCODE)
```

## Dataset 1. Lookup table to link point count station locations (SS) with forest stand polygons

Load intersected file and select only the basic columns required to link a bird point count location (SS) with a given forest stand polygon (FEATURE_ID)
``` {r}
lookup.bird.ss.vri.intersect <- birds.ss[c("PCODE", "SS", "POINT_X", "POINT_Y", "FEATURE_ID")] 
lookup.bird.ss.vri.intersect <- lookup.bird.ss.vri.intersect[order(lookup.bird.ss.vri.intersect$PCODE),]
kable(rbind(head(lookup.bird.ss.vri.intersect),tail(lookup.bird.ss.vri.intersect)))
write.csv(lookup.bird.ss.vri.intersect, file="rawData/LOOKUP.birdSSvriIntersect.csv", row.names=F)
```


## Dataset 2. Bird data table with survey events (date/time) linked to location (SS)

``` {r}
pkey <- read.csv("rawData/BC_COFI_PKEY-manuallyfixed.csv", header=T)
colnames(pkey)
head(pkey)
unique(pkey$PCODE)
```

Check if SS are the same between the dataset intersected with VRI and this PKEY one

``` {r}
all.equal(unique(pkey$SS), unique(lookup.bird.ss.vri.intersect$SS))
```

They're not equal --> pkey file has more. 
``` {r}
testing.equal <- as.character(unique(pkey$SS)) == as.character(unique(lookup.bird.ss.vri.intersect$SS))
head(testing.equal)
head(unique(pkey$SS))
head(unique(lookup.bird.ss.vri.intersect$SS))

```

It's not clear why, but the intersected file seems to be missing some BBS routes from BC. It doesn't matter so much right now, since I'm going to get version 5 of the BAM dataset soon. I'll just continue with whatever points were intersected. 

Merge vri intersected points with pkey data, keeping only those that have been intersected

``` {r}
birds.pkey <- merge(pkey, lookup.bird.ss.vri.intersect, by="SS", all.y=TRUE)
nrow(birds.pkey)
write.csv(birds.pkey, file="rawData/birds.pkey.novri.csv", row.names=F)
```

## Dataset 3: Bird data table with bird observations linked to PKEY and location


## Dataset 4: Bird and vri data table with bird observations linked to VRI data